### 2015-08-28 chapters/conclusion.tex
minor updates

\chapter{Conclusions} \label{chap:conclusion}

One might argue that the space of discovery in Astronomy is spanned by resolution and 
wavelength.  Technology has enabled us to contemplate phenomena at wavelengths far outside 
what the realm of our human senses, revolutionizing fields like high energy astrophysics,
cosmology, and the study of the inter stellar-medium.  Resolution then reveals the hidden
details and helps us to interpret what these new eyes see.  A recent example is 
the application of interferometry at sub-mm wavelengths with ground breaking telescopes like the SMA, 
the amazing new power of ALMA, and the EHT revolution.  
While the science motivating chapters \ref{chap:v4046_structure} 
through \ref{chap:radial_flows} may be different than in the final chapter of this thesis, 
I am left wondering and waiting for what will we see when we look only a little bit closer.

Emboldened by this spirit, we shall conclude not only with summaries of the work presented here,
but also some brief discussion of recent developments in the field.  In chapter \ref{chap:v4046_structure} we 
used the SMA to study the protoplanetary disk 
around the close binary V4046 Sgr.  This disk featured a narrow ring of large mm-sized grains, 
but an extended gas rich disk.  Detailed modeling of these features demonstrated that they were 
phenomenologically consistent with the viscous spreading of the gas and the concentration through pressure 
drag forces of the dust.  This disk is a single example among a number, and more such examples continue to be 
found, \citep{isella07,panic09,andrews12,degregorio-monsalvo13,ke14}.  A somewhat related phenomenon are the 
severely asymmetric dust  ``horse-shoes'' that have been recently observed at multiple wavelengths
\citep{vandermarel13,fukagawa13,isella13,casassus15,marino15}.  Thought to be perhaps generated by gigantic 
3D Rossby wave instabilities \citep{regaly13}, these are another striking example from the interplay between
the gas and dust in protoplanetary disks.

Resolving magnetic fields: \citep{stephens14}

HL Tau: \citep{brogan14}

A gap in Tw Hya at 20 AU? \citep{akiyama15}

Gas density within cavities of transition disks \citep{vandermarel15}

More dust traps: \citep{espaillat15}

\section{A Bright Future}
What is happening in the inner disks? \citep{ke14}
Circumplanetary disks? \citep{perez15}
Chemistry and stuff \citep{oberg15} \citep{graninger15} 
Chemistry as a signpost of planet formation \citep{cleeves15}
Exploring the outer solar system with New Horizons
Detection of organics by Rosetta? Checkout: \url{http://www.sciencedaily.com/releases/2015/07/150730172518.htm}

Summarize status of EHT and its future.

This thesis is a small expression of 
the science that resolution at these wavelengths can provide and only an exploration for the 
hard and promising work ahead.

>>>>

\chapter{Conclusion} \label{chap:conclusion}

We conclude with brief summaries of the preceding chapters along with some discussion of recent developments 
in the field.

In Chapter \ref{chap:v4046_structure} we 
used the SMA to study the protoplanetary disk 
around the close binary V4046 Sgr.  This disk featured a narrow ring of large mm-sized grains along with
an extended gas rich disk.  Detailed modeling of these features demonstrated that they were 
consistent with the viscous spreading of the gas and the concentration of the dust via pressure 
drag forces.  This disk is one published example among a few and more similar systems continue to be
identified \citep{isella07,panic09,andrews12,degregorio-monsalvo13,ke14}.  A related phenomena are the 
severely asymmetric dust  ``horse-shoes'' that have been  observed at multiple wavelengths
\citep{vandermarel13,fukagawa13,isella13,casassus15,marino15}.  Thought to be perhaps generated by gigantic 
3D Rossby wave instabilities \citep{regaly13}, these are another striking example of the interplay between
the gas and dust in protoplanetary disks.  A particular exciting example are the narrow gaps in the dust continuum
of HL Tau (see Figure \ref{fig:hltau}. This striking image was obtained using ALMA baselines as long as 15\,km 
and possible causes include
planet-disk interactions and condensation fronts \citep{brogan14,zhang15}.

\begin{figure}
\plotone{chapters/HLTau_nrao.jpg}
\caption{This ALMA observation of the HL Tau disk revealed multiple rings and gaps \citep{brogan15}.}
\label{fig:hltau}
\end{figure}

In Chapter \ref{chap:hd163296} we explored the vertical structure of a protoplanetary disk using 
observations that exhibited clear signs of the kinetic temperature changing vertically in the disk.
The search for azimuthal temperature variations are a natural extension of this work \citep{isella13} and 
some evidence has been found \citep{vanderplas14}.  There is much activity designing 
observations to probe the ambient conditions 
for planet formation beyond temperature from the local ionization environment \citep{cleeves14,teague15} to 
turbulence \citep{hughes11,guilloteau12,flock15,simon15}.

To conclude our work on protoplanary disks, we discussed in Chapter \ref{chap:radial_flows} observational 
indicators of 
radial gas flows within the holes of transition disks.  \citet{casassus15} have studied the gas kinematics for
the curious case of HD142527 using the CO $J = 6-5$ line and found that their data is consistent with
accretion reaching close to free-fall velocities.  The gas surface density of transition disks have also 
been recently studied using observations taking at sub-millimeter wavelengths 
\citep{bruderer14,zhang14,canovas15,perez15,vandermarel15} and will help determine how these structures are formed
\citep{bruderer13}.  Molecular disk winds, outflows, and binary systems are all recently observed 
with interesting gas kinematics \citep{klaasen13,dutrey14,williams14,salyk14,czekala15}.  Looking to the 
future, well resolved, highly sensitive observations of spectral lines may constrain the strength of MRI-drive
turbulence \citep{simon15}, query the active sites of planet formation \citep{kleeves15,ober15}, or reveal
circumplanetary disks \citep{perez15}.

The ground-breaking Event Horizon Telescope continues to report exciting new science with every
campaign from the early detections of AGN at mm wavelengths \citep{doeleman05} to structures
at the event horizon of a supermassive black hole \citep{doeleman08,doeleman09,doeleman13}.
The 2015 campaign represented a new scale of operation, incorporating more telescopes and
technologies.  In Chapter \ref{chap:aphids} we reported on the status of the APHIDS software which
will post-process the phased SMA data and make the sampling rate compatible with the rest of the EHT
network.  This problem is not limited to the SMA and we hope that the development done with APHIDS will
prove useful for other sites and VLBI experiments.

This thesis explores
ence that resolution at these wavelengths can provide and only an exploration for the 
hard and promising work ahead.

### 2015-08-28 chapters/introduction.tex
minor updates

Over the past several decades, observations from interferometers operating at millimeter wavelengths have made
significant contributions to understanding the structure of these disks.  The early maps of the CO emission
made using 3 to 5 telescopes in an array at Owens Valley revealed bound, rotating disks with diameters 
similar to the dimensions of own solar system \citep{sargent87,weintraub89,koerner93}.  Emission from other 
molecules were then surveyed \citep{dutrey97} and studies of multiple isotopologues were used to constrain the 
two-dimensional structure of some disks \citep{dartois03,pietu07,qi11} aided by the development of 
chemical models \citep{aikawa99,aikawa01}.  Images of the dust also revealed interesting structures such as
giant holes centered around the star for a class of ``transition disks'' \citep{hughes07,brown09,andrews11} and 
evidence for binary interactions \citep{akeson98}.

@@

The main two ingredients in the disk are gas and dust.  The majority of the disk mass will be in molecular 
hydrogen, but observations focus on easier to detect trace molecules such as $^{12}$CO and its isotopologues.
The dust grains are detected via their bulk thermal emission which, when the emisison is optically thin,  
can give a measure of the dust mass \citep{hildebrand83}.  Interferometric observations taken at 
millimeter wavelengths can resolve structure of nearby protoplanetary disks on scales of a few tens to 
hundreds of astronomical units.   The bulk motion of material in the disk can also be 
studied using these same telescopes by studying how the pattern of line emission
from molecules such as CO changes with frequency \citep{beckwith93}.  The velocity of the gas relative to the 
observer 
changes as a function of position in the disk. To an observer scanning over frequency, different parts of the 
disk seem to light up when its Doppler shifted emission matches the observer's frequency window.

There exists several techniques for constraining the structure of protoplanetary disks.  Some geometric 
properties such as size, inclination, and orientation can be calculated from an image.  The amount of dust 
in a disk can also be estimated by assuming a dust opacity and measuring the mm-wave flux.  Alternatively, one 
may try to solve the inverse problem: assume some disk model, calculate the resulting emission, and then compare
it to what is observed.  Often the density, temperature, and velocity profiles of the disks are assumed to 
follow power laws or some other parametric function.  There also exist some grids and models that try to 
calculate abundances in a self-consistent manner.  Additionally, researchers usually either try to constrain 
populations using 
surveys \citep[i.e.][]{andrews11,oberg11,harris12} or present detailed studies of single, 
interesting disks.  The work
presented in chapters \ref{chap:v4046sgr_structure} to \ref{chap:radial_flows} takes the latter approach and uses simple parametric 
models to investigate the structure of a choice few, but very interesting, protoplanetary disks.

V4046 Sgr is a system with two stars closely orbiting each other.  Around this tight binary is a bright
protoplanetary disk.  In chapter \ref{chap:v4046sgr_structure} we present SMA observations 
of this disk that reveal a large central cavity in the dust.  In addition, we show that the distribution dust 
does not follow that of the gas: the gas disk is large and extended while the 
mm-sized dust grains are concentrated in a narrow ring.  Furthermore, it appears that the amount of dust left 
within the ring along with the size of the canvity is inconsistent with photoevaporative scenarios.  We argue 
that this configuration is the hallmark of an old disk that is being sculpted by at least one embedded planet.

HD163296 also has a very large, gas rich disk, but no observed central cavity.  In chapter \ref{chap:hd163296} 
we use science verification data from ALMA to study the vertical structure of 
this disk.  We argue that resolved features in the images highlight how the temperature of the gas changes with
its distance from mid-plane and explore how additional, nuanced aspects of the disk structure may affect
the data.  Chapter \ref{chap:radial_flows} then considers what impact radial motions of gas within transition 
disk holes may have upon spectral line emission.  We discuss several indicators of such radial flows along 
with possible sources of confusion such as warps in the disk.  One disk has provided intriguing
evidence of such flows \citep{casassus13,cassasus15} and future observations may provide others.

@@

The Event Horizon Telescope (EHT) observes supermassive black holes, obtaining resolution high enough
to resolve event horizon scales of nearby targets.  These observations provide a 
unique glimpse into one of the most extreme environments in our universe.  Using telescopes in Hawaii, Arizona, 
and California, the collaboration has already detected structure at these scales for Sgr A$^\ast$ in the galactic 
center and the giant elliptical galaxy M87 \citep{doeleman08,doeleman09}.  These data have provided indirect 
evidence for existence of an event horizon \citep{broderick15} along with constraints on the degree of 
magnetization in the jet of M87 \citep{kino15} and the orientation of Sgr A$^\ast$ \citep{broderick11}.  EHT 
observations promise to provide tests of general relativity \citep{luminet79,johannsen10,bambi13,broderick14} as 
well as new studies of black hole time variability \citep{doeleman09} and black hole accretion \citep{chan15}.

The bright future of the EHT rests on its key ability to expand by adding more telescopes and
incorporating new technologies into its operations.  The March 2015 observing campaign was a major milestone in
the EHT expansion with participation from the Large Millimeter Telescope (LMT) in Mexico, Combined Array for 
Research in Millimeter-wave Astornome (CARMA) in California, the SMA and James Clerk Maxwell Telescope (JCMT) in
 Hawaii, SMT in Arizona and the 30m dish at Pico Valeta in Spain.  However, the data taken at the SMA must be processed before it
can be correlated with the rest of the array.  In chapter \ref{chap:aphids} we describe the APHIDS pipeline
which performs a sampling rate conversion using a cluster of Graphcs Processing Units (GPUs).  This software
is intended to be a long-term solution to enable VLBI with the SMA and has applications for other sites such 
as ALMA that operate at a sampling rate that is not compatible with the standard VLBI data product.

>>>>

Over the past several decades, observations from interferometers operating at millimeter wavelengths have made
significant contributions to understanding the structure of these disks.  The early maps of the CO emission
made using 3 to 5 telescopes in an array at Owens Valley revealed bound, rotating disks with diameters 
similar to the dimensions of own solar system \citep{sargent87,weintraub89,koerner93}.  Emission from other 
molecules were then surveyed \citep{dutrey97} and studies of multiple isotopologues were used to constrain the 
two-dimensional structure of some disks \citep{dartois03,pietu07,qi11} aided by the development of 
chemical models \citep{aikawa99,aikawa01}.  Images of the dust also revealed interesting structures such as
giant holes centered around the star for a class of ``transition disks'' \citep{hughes07,brown09,andrews11} and 
evidence for binary interactions \citep[i.e.][]{akeson98}.

@@

The main two ingredients in the disk are gas and dust.  The majority of the disk mass will be in molecular 
hydrogen, but observations focus on easier to detect trace molecules such as $^{12}$CO and its isotopologues.
The dust grains are detected via their bulk thermal emission which, when the emisison is optically thin,  
can be used to measure of dust mass \citep{hildebrand83}.  Interferometric observations taken at 
millimeter wavelengths can resolve structure of nearby protoplanetary disks on scales of a few tens to 
hundreds of astronomical units.  The bulk motion of material in the disk can also be 
studied using these same telescopes by studying how the pattern of line emission
from molecules such as CO changes with frequency \citep{beckwith93}.  The velocity of the gas relative to the 
observer changes as a function of position in the disk. To an observer scanning over frequency, different parts 
of the disk seem to light up when its Doppler shifted emission matches the observer's frequency window.

There exists several techniques for constraining the structure of protoplanetary disks.  Some geometric 
properties such as size, inclination, and orientation can be calculated from an image.  The amount of dust 
in a disk can also be estimated by assuming a dust opacity and measuring the millimeter wave flux.  
Alternatively, one may try to solve the inverse problem: assume some disk model, calculate the resulting 
emission, and then compare the model's prediction to what is observed \citep[i.e.][]{dartois03,qi03}.  Often the 
density, temperature, and velocity profiles of the disks are assumed to 
follow power laws or some other parametric function.  There also exist some grids and models that try to 
calculate abundances in a self-consistent manner.  Additionally, researchers usually either try to constrain 
populations using 
surveys \citep[i.e.][]{andrews11,oberg11,harris12} or present detailed studies of single, 
interesting disks.  The work
presented in chapters \ref{chap:v4046sgr_structure} to \ref{chap:radial_flows} takes the latter approach and 
uses simple parametric 
models to investigate the structure of a choice few, but very interesting, protoplanetary disks.

V4046 Sgr is a system with two stars closely orbiting each other and around this tight binary is a bright
protoplanetary disk.  In chapter \ref{chap:v4046sgr_structure} we present SMA observations 
of this disk that reveal a large central cavity in the dust.  In addition, we show that the distribution dust 
does not follow that of the gas: the gas disk is large and extended while the 
millimeter sized dust grains are concentrated in a narrow ring.  Furthermore, thethe amount of dust left inside
the cavity is inconsistent with photoevaporative scenarios.  We argue 
that this configuration is the hallmark of an old disk that is being sculpted by at least one unobserved 
companion.

HD163296 also has a very large, gas rich disk, but no observed central cavity.  In chapter \ref{chap:hd163296} 
we use science verification data from ALMA to study the vertical structure of 
this disk.  We argue that resolved features in the images highlight how the temperature of the gas changes with
its distance from the disk mid-plane and explore how additional, nuanced aspects of the disk structure may affect
the data.  Chapter \ref{chap:radial_flows} then considers what impact radial motions of gas within transition 
disk holes may have upon spectral line emission.  We discuss several indicators of such radial flows along 
with possible sources of confusion such as warps in the disk.  One disk has provided intriguing
evidence of such flows \citep{casassus13,cassasus15} and future observations may provide others.

@@

The Event Horizon Telescope (EHT) observes supermassive black holes with baselines long enough to
to resolve event horizon scales of nearby targets.  These VLBI observations provide a 
unique glimpse into one of the most extreme environments in our universe.  Using telescopes in Hawaii, Arizona, 
and California, the collaboration has already detected structure at these scales for Sgr A$^\ast$ in the galactic 
center and the blaock hole at the center of the giant elliptical galaxy M87 
\citep{doeleman08,doeleman09}.  These data have provided indirect 
evidence for the existence of an event horizon \citep{broderick15} along with constraints on the degree of 
magnetization in the jet of M87 \citep{kino15} and the orientation of Sgr A$^\ast$ \citep{broderick11}.  EHT 
observations promise to provide tests of general relativity \citep{luminet79,johannsen10,bambi13,broderick14} as 
well as new studies of black hole time variability \citep{doeleman09} and accretion \citep{chan15}.

The bright future of the EHT is partly driven by its ability to add more telescopes and
incorporate new technologies.  The March 2015 observing campaign was a major milestone in
the EHT expansion with participation from the Large Millimeter Telescope (LMT) in Mexico, Combined Array for 
Research in Millimeter-wave Astornomy (CARMA) in California, the SMA and James Clerk Maxwell Telescope (JCMT) in
 Hawaii, SMT in Arizona and the 30m dish at Pico Valeta in Spain.  However, the data taken at the SMA must be 
processed before it
can be correlated with the rest of the array.  In chapter \ref{chap:aphids} we describe the APHIDS pipeline
which performs a sampling rate conversion using a cluster of Graphcs Processing Units (GPUs).  This software
is intended to be a long-term solution enabling VLBI with the SMA and has applications for other sites such 
as ALMA.
